\documentclass[12pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{geometry}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{array}
\usepackage{doi}

\geometry{margin=2.5cm}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue
}

\title{Gauge--Gravity Spectral Synthesis:\\
Joint Einstein--Yang--Mills Equations, Born--Infeld Completion,\\
and the Hierarchy Ratio from Projective Spectral Entropy}

\author{J.\ Beau\\
\small Independent Researcher, France\\
\small \texttt{jerome.beau@cosmochrony.org}}

\date{2026}

\begin{document}

  \maketitle

  \begin{abstract}
    The companion papers Gravity~3.0~\cite{Beau2026h} and Q12~\cite{Beau2026q12}
    derived, respectively, the Einstein tensor as the horizontal $a_2$ response
    and the Yang--Mills equations as the vertical $a_4$ response of the projective
    spectral entropy functional $S_\Pi[g,A] = \tfrac{1}{2}\log\det' \mathcal{A}_{g,A}$.
    The present paper carries out the gauge--gravity synthesis and establishes its
    central conceptual consequence: \emph{gravity and gauge dynamics are not unified
    by enlarging the symmetry group or adding fields, but are stratified by the
    Seeley--DeWitt level at which the same admissible operator responds to variation.}
    The conventional question ``what symmetry unifies gravity and gauge?'' is thereby
    replaced by ``at what spectral order does the projection respond?''

    We solve the \emph{joint} variational problem $\delta_{g,A}S_\Pi = 0$ and
    derive the coupled Einstein--Yang--Mills system: the Yang--Mills stress tensor
    sources gravity through the metric dependence of the $a_4$ coefficient, with
    coupling constant $8\pi G_N = c_{YM}/c_{EH}$ fixed by the Seeley--DeWitt expansion
    alone.
    We compute the $a_6$ coefficient and identify the leading gauge--gravity
    cross-coupling $R_{\mu\nu\rho\sigma}F^{\mu\nu}F^{\rho\sigma}$.
    The Eddington--Born--Infeld completion of the joint functional is constructed
    from the same Born--Infeld saturation bound that governs the gravitational
    sector~\cite{Beau2026c}.
    The gauge--gravity hierarchy is a structural output of the spectral stratification:
    gravity is induced at $a_2$ (quadratic UV divergence) and gauge at $a_4$
    (logarithmic UV divergence), from the same cutoff $\ell_{\mathrm{sp}}$, giving:
    \[
      G_N g^2_{YM}
      \;\sim\;
      \frac{\ell_{\mathrm{sp}}^2}{\dim V}
      \cdot\Bigl[\log\frac{\Lambda}{\mu}\Bigr]^{-1}
      \;\ll\; 1,
    \]
    without fine-tuning.
    The spectral stratification
    $a_2 \to \text{gravity},\; a_4 \to \text{gauge},\; a_6 \to \text{mixed coupling}$
    predicts that fermionic structure should appear at a dedicated spectral level
    carrying a Dirac-type admissible operator, which is the central open question
    for the continuation of the programme.
    The gauge group $G_\Pi$ is inherited from the spectral admissibility
    sub-programme~\cite{Beau2026q6a,Beau2026a35}; the $\mathrm{SU}(3)$ sector is unconditional:
    $[H\text{-color}]_{\mathrm{eff}}$ (equality of capacity exponents) and
    $[H\text{-color}]_{\mathrm{pointwise}}$ (exact equality at finite~$q$)
    are both established analytically in~\cite{Beau2026a35,Beau2026a36}.
  \end{abstract}

  \textbf{Keywords:} gauge--gravity synthesis, Einstein--Yang--Mills equations,
  Seeley--DeWitt expansion, spectral entropy functional, Eddington--Born--Infeld
  completion, gauge--gravity hierarchy, non-injective projection, admissible bundle.

  \tableofcontents

  \clearpage
  \section{Introduction}
    \label{sec:intro}

    \subsection{Context}
      \label{ssec:context}

      The Cosmochrony framework derives emergent physics from a generally non-injective
      projection $\Pi\colon\chi\to\mathcal{M}$ of a static relational substrate $\chi$.
      Within this framework, the Q-series of papers establishes the emergence of
      quantum structure, spacetime geometry, and gauge dynamics from the admissibility
      structure of $\Pi$~\cite{Beau2026m,Beau2026q1,Beau2026q2}.
      The spectral entropy functional
      \[
        S_\Pi[g] = \tfrac{1}{2}\log\det'\ \mathcal{A}_g,
      \]
      where $\mathcal{A}_g = -(\nabla^g)^2$ is the admissible Laplacian, encodes
      the effective geometric dynamics of the projection.
      Its metric variation produces the Einstein tensor as the infrared-dominant
      local response via the $a_2$ Seeley--DeWitt coefficient~\cite{Beau2026h}.

      The Q12 paper~\cite{Beau2026q12} extended the functional to the gauge sector
      by coupling $\mathcal{A}_g$ to an admissible connection $A$ on the associated
      vector bundle:
      \[
        S_\Pi[g,A] = \tfrac{1}{2}\log\det'\ \mathcal{A}_{g,A},
        \qquad
        \mathcal{A}_{g,A} = -(\nabla^A)^2 + E.
      \]
      The vertical variation $\delta_A S_\Pi = 0$ then yields the Yang--Mills equations
      $D_\mu F^{a\mu\nu} = 0$ through the $a_4$ coefficient.
      The two results place gravity and Yang--Mills dynamics at different levels of
      the Seeley--DeWitt hierarchy of the \emph{same} functional, separated by geometric
      direction (horizontal versus vertical variation) and by spectral order.

    \subsection{Main results}
      \label{ssec:main}

      The present paper delivers the four items deferred from Q12~\cite{Beau2026q12}:

      \begin{enumerate}
        \item \textbf{Joint variational problem.}
        We solve $\delta_{g,A}S_\Pi = 0$ as a coupled system and show that the
        Yang--Mills stress tensor $T^{\mathrm{YM}}_{\mu\nu}$ enters the metric equation
        as the unique source term produced by the $a_4$ coefficient under metric variation
        (Theorem~\ref{thm:EYM}).

        \item \textbf{Gauge--gravity coupling at order $a_6$.}
        We compute the leading cross-term in the $a_6$ Seeley--DeWitt coefficient,
        which contains $R_{\mu\nu\rho\sigma} F^{\mu\nu} F^{\rho\sigma}$ and encodes
        the gravitational backreaction of the gauge field at higher spectral order
        (Theorem~\ref{thm:a6}).

        \item \textbf{Eddington--Born--Infeld completion.}
        We construct the EBI extension of $S_\Pi[g,A]$ from the same saturation
        bound $c_\chi$ that underlies the gravitational Born--Infeld completion
        of~\cite{Beau2026c} (Theorem~\ref{thm:EBI}).

        \item \textbf{Quantitative hierarchy ratio.}
        We derive a precise structural bound on $G_N g^2_{YM}$ and show that it
        is structurally suppressed in both $\ell_{\mathrm{sp}}^2$ and $[\log(\Lambda/\mu)]^{-1}$, without fine-tuning
        (Proposition~\ref{prop:hierarchy}).
      \end{enumerate}

    \subsection{Organisation}
      \label{ssec:org}

      Section~\ref{sec:joint} states the joint functional and recalls the
      horizontal--vertical decoupling structure.
      Section~\ref{sec:EYM} derives the Einstein--Yang--Mills system.
      Section~\ref{sec:a6} computes the $a_6$ cross-coupling.
      Section~\ref{sec:EBI} constructs the Eddington--Born--Infeld completion.
      Section~\ref{sec:hierarchy} derives the hierarchy ratio quantitatively.
      Section~\ref{sec:epistemic} discusses the epistemic status and analytical
      status of [H-color].
      Section~\ref{sec:conclusion} concludes.

      \clearpage
  \section{The joint spectral entropy functional}
    \label{sec:joint}

    \subsection{The admissible principal bundle}
      \label{ssec:bundle}

      Following Q12~\cite{Beau2026q12}, let $(M,g)$ be a compact Riemannian four-manifold
      and let $P\to M$ be a principal $G_\Pi$-bundle, where $G_\Pi$ is the admissible gauge
      group inherited from the spectral admissibility sub-programme~\cite{Beau2026q6a}.
      Let $E = P\times_\rho V$ be the associated vector bundle for a representation
      $\rho\colon G_\Pi\to \mathrm{GL}(V)$.
      An admissible connection $A\in\mathcal{A}(P)$ defines a covariant derivative
      $\nabla^A$ on $E$ and a curvature two-form $F_A = dA + A\wedge A\in\Omega^2(M,\mathrm{ad}\,P)$.

    \subsection{The Laplace-type operator and its decomposition}
      \label{ssec:Laplace}

      Define the Laplace-type operator on sections of $E$:
      \[
        \mathcal{A}_{g,A} = -(\nabla^A)^2 + E,
      \]
      where $E\in\Gamma(\mathrm{End}\,E)$ is the endomorphism determined by the admissibility
      structure.
      By the horizontal--vertical decoupling lemma of Q12 (Lemma~1 of~\cite{Beau2026q12}),
      the operator splits as:
      \[
        \mathcal{A}_{g,A} = \mathcal{A}_g^{\mathrm{hor}} \otimes \mathrm{Id}_V
        \;+\;
        \mathrm{Id}_{\mathcal{H}} \otimes \mathcal{A}_A^{\mathrm{vert}}
        \;+\;
        \mathcal{A}^{\mathrm{mix}},
      \]
      where $\mathcal{A}^{\mathrm{mix}}$ is suppressed at the spectral scale $\ell_{\mathrm{sp}}$
      by minimal coupling.
      The spectral entropy functional is then:
      \[
        S_\Pi[g,A] = \tfrac{1}{2}\mathrm{Tr}\log\mathcal{A}_{g,A}
        \;=\;
        \tfrac{1}{2}\log\det'\,\mathcal{A}_{g,A}.
      \]
      This is the common generating functional for both the gravitational and gauge sectors.

    \subsection{Seeley--DeWitt expansion of the joint functional}
      \label{ssec:SDW}

      The heat-kernel expansion of $\mathcal{A}_{g,A}$ gives~\cite{Vassilevich2003,ParkerToms2009}:
      \begin{equation}
        \mathrm{Tr}\,e^{-t\mathcal{A}_{g,A}}
        \;\sim\;
        \sum_{k=0}^\infty a_{2k}[\mathcal{A}_{g,A}]\,t^{(2k-4)/2}
        \quad (t\to 0^+),
        \label{eq:heat-kernel}
      \end{equation}
      where the Seeley--DeWitt coefficients $a_{2k}$ are local integrals of
      $g$, $A$, and their covariant derivatives.
      After renormalization of the UV-divergent terms, the functional decomposes as:
      \begin{align}
        S_\Pi[g,A]
        &= c_{(0)}\Lambda^4 a_0
        \;+\;
        c_{(2)}\Lambda^2 a_2[g]
        \;+\;
        c_{(4)}\log\frac{\Lambda}{\mu}\,a_4[g,A]
        \;+\;
        S_\Pi^{\mathrm{fin}}[g,A]
        \;+\;
        O\!\left(\frac{1}{\Lambda^2}\right),
        \label{eq:SDW-expansion}
      \end{align}
      where $S_\Pi^{\mathrm{fin}}$ is the finite part at the renormalization scale $\mu$.
      The cosmological-constant sector $a_0$ does not contribute to the equations of motion
      and is set aside.
      The two dynamically active sectors are:
      \begin{align}
        a_2[g]
        &= \frac{1}{16\pi^2}\int_M \frac{R}{6}\,\mathrm{tr}(\mathrm{Id}_V)\,\sqrt{g}\,d^4x,
        \label{eq:a2}
        \\[4pt]
        a_4[g,A]
        &= \frac{1}{16\pi^2}\int_M
        \Bigl[
          \alpha_1 R^2 + \alpha_2 R_{\mu\nu}R^{\mu\nu}
          + \alpha_3 R_{\mu\nu\rho\sigma}R^{\mu\nu\rho\sigma}
          + \tfrac{1}{12}\mathrm{Tr}(\Omega^2_A)
          \Bigr]\sqrt{g}\,d^4x,
        \label{eq:a4}
      \end{align}
      where $\Omega_A^{\mu\nu} = F^a_{\mu\nu}\,T^a$ is the curvature two-form in the
      representation $\rho$ and $\alpha_1,\alpha_2,\alpha_3$ are numerical coefficients
      depending on the spectrum of $\mathcal{A}_g$~\cite{Vassilevich2003}.
      The Yang--Mills term $\tfrac{1}{12}\mathrm{Tr}(\Omega^2_A)$ is the unique
      $a_4$ contribution that is first-order in the curvature two-form.

  \section{Joint Einstein--Yang--Mills equations of motion}
    \label{sec:EYM}

    \subsection{Metric variation of the joint functional}
      \label{ssec:metric-var}

      The metric variation of $S_\Pi[g,A]$ receives contributions from all
      Seeley--DeWitt coefficients that depend on $g$.
      At leading infrared order, the dominant term is:
      \begin{equation}
        \frac{\delta S_\Pi}{\delta g^{\mu\nu}}
        \;=\;
        c_{EH}\,G_{\mu\nu}\sqrt{g}
        \;+\;
        c_{YM}\,\frac{\delta}{\delta g^{\mu\nu}}
        \int_M \mathrm{Tr}(\Omega^2_A)\sqrt{g}\,d^4x
        \;+\;
        O(R^3,\,R\cdot F^4),
        \label{eq:metric-var}
      \end{equation}
      where $c_{EH} = c_{(2)}\Lambda^2/(96\pi^2)$ and $c_{YM} = c_{(4)}\log(\Lambda/\mu)/(192\pi^2)$
      are the induced coupling constants established in~\cite{Beau2026h} and~\cite{Beau2026q12} respectively.

      \begin{lemma}[Metric variation of the Yang--Mills term]
        \label{lem:YM-stress}
        Let $\mathcal{L}_{YM} = -\tfrac{1}{4}\mathrm{Tr}(F_{\mu\nu}F^{\mu\nu})\sqrt{g}$.
        Then:
        \[
          \frac{1}{\sqrt{g}}\frac{\delta}{\delta g^{\mu\nu}}
          \int_M \mathrm{Tr}(\Omega^2_A)\sqrt{g}\,d^4x
          \;=\;
          -4\,\mathrm{Tr}(F_{\mu\rho}F_\nu{}^\rho)
          \;+\;
          g_{\mu\nu}\,\mathrm{Tr}(F_{\rho\sigma}F^{\rho\sigma})
          \;\equiv\;
          T^{YM}_{\mu\nu}.
        \]
      \end{lemma}

      \begin{proof}
        Standard variation of the Yang--Mills kinetic term with respect to the metric,
        using $\partial_\alpha\sqrt{g} = \tfrac{1}{2}\sqrt{g}\,g^{\mu\nu}\partial_\alpha g_{\mu\nu}$
        and the fact that $F_{\mu\nu}$ is metric-independent.
        The result is the standard Yang--Mills energy-momentum tensor.
      \end{proof}

    \subsection{Gauge-field variation}
      \label{ssec:gauge-var}

      From Theorem~1 of Q12~\cite{Beau2026q12}, the variation of $S_\Pi[g,A]$ with
      respect to the admissible connection $A$ yields:
      \[
        \frac{\delta S_\Pi}{\delta A^a_\nu}
        \;=\;
        c_{YM}\,D_\mu F^{a\mu\nu}\sqrt{g}
        \;+\;
        O(F^3,\,R\cdot F),
      \]
      where $D_\mu = \partial_\mu + [A_\mu,\,\cdot\,]$ is the gauge-covariant derivative.
      The Yang--Mills equations $D_\mu F^{a\mu\nu} = 0$ follow at leading order.

    \subsection{The joint system}
      \label{ssec:joint}

      \begin{theorem}[Joint Einstein--Yang--Mills equations]
        \label{thm:EYM}
        Let $S_\Pi[g,A] = \tfrac{1}{2}\log\det'\,\mathcal{A}_{g,A}$ be the joint
        projective spectral entropy functional.
        The joint variational problem $\delta_{g,A}S_\Pi = 0$, at leading order in
        the Seeley--DeWitt expansion, is equivalent to:
        \begin{align}
          G_{\mu\nu}
          &= 8\pi G_N\,T^{YM}_{\mu\nu},
          \label{eq:Einstein-YM}
          \\[4pt]
          D_\mu F^{a\mu\nu}
          &= 0,
          \label{eq:YM}
        \end{align}
        where the coupling constant is defined by:
        \[
          8\pi G_N \;=\; \frac{c_{YM}}{c_{EH}}.
        \]
      \end{theorem}

      \begin{proof}
        Setting $\delta_g S_\Pi = 0$ using~\eqref{eq:metric-var} and Lemma~\ref{lem:YM-stress}:
        \[
          c_{EH}\,G_{\mu\nu} + c_{YM}\,T^{YM}_{\mu\nu} = 0.
        \]
        The standard form~\eqref{eq:Einstein-YM} follows by identifying $8\pi G_N = c_{YM}/c_{EH}$.
        Consistency with the explicit induced values established in~\cite{Beau2026h,Beau2026q12}:
        \begin{equation}
          G_N^{-1} \;=\; \frac{\dim V}{16\pi^2}\,\ell_{\mathrm{sp}}^{-2},
          \qquad
          g_{YM}^{-2} \;=\; \frac{1}{12\cdot 16\pi^2}\,\log\frac{\Lambda}{\mu},
          \label{eq:induced-couplings}
        \end{equation}
        follows from $c_{EH} \propto \dim V\,\Lambda^2$ and $c_{YM} \propto \log(\Lambda/\mu)$
        with $\Lambda \sim \ell_{\mathrm{sp}}^{-1}$:
        the ratio $8\pi G_N = c_{YM}/c_{EH} \propto \ell_{\mathrm{sp}}^2\,\log(\Lambda/\mu)/\dim V$
        grows with $\log\Lambda$ at fixed $\ell_{\mathrm{sp}}$, while $G_N^{-1} \sim \ell_{\mathrm{sp}}^{-2}$
        is the dominant quadratic contribution.
        In particular $G_N \sim \ell_{\mathrm{sp}}^2$ (not $\ell_{\mathrm{sp}}^{-2}$).
        Equation~\eqref{eq:YM} is Theorem~1 of~\cite{Beau2026q12} and holds unchanged
        since $\delta_A S_\Pi$ does not involve the $a_2$ sector.
      \end{proof}

      \begin{remark}
        The sign in~\eqref{eq:Einstein-YM} reflects the Euclidean signature convention
        of the spectral derivation: the Yang--Mills energy density is positive, and the
        Yang--Mills stress tensor acts as a positive source of curvature, as expected.
        The coupling constant $8\pi G_N$ is \emph{not an input} but is determined by
        the ratio of Seeley--DeWitt coefficients, both of which are fixed by the same
        spectral cutoff $\ell_{\mathrm{sp}}$.
      \end{remark}

  \section{Gauge--gravity coupling at order \texorpdfstring{$a_6$}{a6}}
    \label{sec:a6}

    \subsection{The \texorpdfstring{$a_6$}{a6} Seeley--DeWitt coefficient}
      \label{ssec:a6-coeff}

      The $a_6$ coefficient of the operator $\mathcal{A}_{g,A}$ is a local integral
      of terms of total derivative order six.
      For a Laplace-type operator on a vector bundle with curvature endomorphism $E$
      and connection curvature $\Omega_A$, the general form
      is~\cite{Vassilevich2003,ParkerToms2009}:
      \begin{align}
        a_6[\mathcal{A}_{g,A}]
        &= \frac{1}{16\pi^2}\int_M
        \Bigl[
          \beta_1 R^3
          + \beta_2 R\,R_{\mu\nu}R^{\mu\nu}
          + \beta_3 R_{\mu\nu}{}^{\rho\sigma}R_{\rho\sigma}{}^{\alpha\beta}
          R_{\alpha\beta}{}^{\mu\nu}
          \nonumber\\[2pt]
          &\qquad
          + \gamma_1\,R_{\mu\nu\rho\sigma}\,\mathrm{Tr}(\Omega^{\mu\nu}_A\,\Omega^{\rho\sigma}_A)
          + \gamma_2\,R\,\mathrm{Tr}(\Omega^2_A)
          + \gamma_3\,\mathrm{Tr}(\Omega^4_A)
          + \gamma_4\,\mathrm{Tr}((\nabla\Omega_A)^2)
          \Bigr]\sqrt{g}\,d^4x,
        \label{eq:a6}
      \end{align}
      where $\beta_i$ and $\gamma_i$ are numerical coefficients computable from the
      Seeley--DeWitt algorithm~\cite{Vassilevich2003}.

    \subsection{The leading cross-coupling term}
      \label{ssec:cross}

      The dynamically significant cross-coupling between gravity and gauge at order $a_6$
      is encoded in the $\gamma_1$ term:
      \begin{equation}
        \mathcal{L}^{(6)}_{\mathrm{cross}}
        = \frac{\gamma_1}{16\pi^2}\,
        R_{\mu\nu\rho\sigma}\,\mathrm{Tr}(F^{\mu\nu}\,F^{\rho\sigma})\,\sqrt{g}.
        \label{eq:cross-coupling}
      \end{equation}
      This term couples the Riemann tensor directly to the gauge curvature two-form
      and produces corrections to both the Einstein equation and the Yang--Mills equation:

      \begin{theorem}[$a_6$ cross-coupling]
        \label{thm:a6}
        At order $a_6$, the joint equations of motion receive the corrections:
        \begin{align}
          G_{\mu\nu}
          &= 8\pi G_N\,T^{YM}_{\mu\nu}
          \;+\;
          c_{(6)}\log\frac{\Lambda}{\mu}
          \Bigl[
            \nabla^\rho\nabla^\sigma\mathrm{Tr}(F_{\mu\rho}F_{\nu\sigma})
            + \text{(curvature terms)}
            \Bigr]
          + O(a_8),
          \label{eq:Einstein-YM-a6}
          \\[4pt]
          D_\mu F^{a\mu\nu}
          &= c_{(6)}'\log\frac{\Lambda}{\mu}\,
          [R^{\mu\nu\rho\sigma},F_{\rho\sigma}]^a
          + O(F^3,\,R^2 F),
          \label{eq:YM-a6}
        \end{align}
        where $c_{(6)}, c_{(6)}' > 0$ are scheme-dependent positive constants of order
        $\ell_{\mathrm{sp}}^2/16\pi^2$.
      \end{theorem}

      \begin{proof}
        The corrections follow from the variation of the $a_6$ term~\eqref{eq:a6}
        with respect to $g^{\mu\nu}$ and $A^a_\nu$ respectively.
        The metric variation of the $\gamma_1$ term uses the identity
        \[
          \frac{\delta}{\delta g^{\mu\nu}}
          \int R_{\alpha\beta\rho\sigma}\,\mathrm{Tr}(F^{\alpha\beta}F^{\rho\sigma})\,\sqrt{g}\,d^4x
          \ni
          2\,\nabla^\alpha\nabla^\beta\mathrm{Tr}(F_{\mu\alpha}F_{\nu\beta})
          + \text{(curvature contractions)},
        \]
        obtained by integration by parts of the Palatini variation.
        The connection variation of the $\gamma_1$ term gives the commutator structure
        in~\eqref{eq:YM-a6} through $\delta\Omega^{\mu\nu}_A = D^\mu(\delta A^\nu) - D^\nu(\delta A^\mu)$.
        The constants $c_{(6)}, c_{(6)}'$ are suppressed by two additional powers of
        $\ell_{\mathrm{sp}}$ relative to $c_{EH}$ and $c_{YM}$, confirming the
        Seeley--DeWitt order counting.
      \end{proof}

      \begin{remark}
        The corrections in Theorem~\ref{thm:a6} are suppressed by $\ell_{\mathrm{sp}}^2$
        relative to the leading-order terms and are therefore negligible at scales
        $\ell \gg \ell_{\mathrm{sp}}$.
        At scales approaching $\ell_{\mathrm{sp}}$, the $a_6$ terms become comparable
        to the Born--Infeld saturation corrections and the EBI completion of
        Section~\ref{sec:EBI} provides the appropriate non-linear resummation.
      \end{remark}

  \section{Eddington--Born--Infeld completion}
    \label{sec:EBI}

    \subsection{Motivation and strategy}
      \label{ssec:EBI-motivation}

      The induced gravity functional $c_{EH} a_2[g]$ diverges quadratically as
      $\Lambda\to\infty$, and the induced gauge functional $c_{YM} a_4[g,A]$ diverges
      logarithmically.
      In both sectors, the Born--Infeld saturation bound $c_\chi$ of the admissibility
      structure~\cite{Beau2026c} provides a natural non-linear completion that regulates
      these divergences at the spectral length $\ell_{\mathrm{sp}}$.
      The gravitational EBI completion was established in~\cite{Beau2026c,Beau2026h};
      the present section constructs the gauge-sector completion and then assembles
      the joint EBI functional.

    \subsection{Gauge-sector Born--Infeld completion}
      \label{ssec:gauge-BI}

      In the gauge sector, the Born--Infeld saturation condition requires that the
      effective amplitude of the gauge field satisfies $A_n \leq c_\chi / \sqrt{\lambda_n}$
      at each spectral mode~\cite{Beau2026c}.
      At the level of the field theory, the natural BI extension involves the
      combination $\delta^\mu_\nu + \ell_{\mathrm{sp}}^2 F^\mu{}_\nu$; however,
      for a non-abelian $F^a_{\mu\nu}$, the antisymmetry $F_{\mu\nu} = -F_{\nu\mu}$
      implies that $\sqrt{\det(\delta + \ell^2_{\mathrm{sp}} F)}$ is not automatically
      real (it equals the Pfaffian of $F$, which can change sign in Euclidean signature).
      The unambiguous non-abelian replacement uses the symmetric positive-semidefinite
      combination $M_{\mu\nu} \equiv F_{\mu\rho}F^\rho{}_\nu$, which is gauge-covariant
      and has non-negative eigenvalues.
      The gauge BI Lagrangian is therefore written as:
      \begin{equation}
        \mathcal{L}_{BI}[A]
        = \frac{1}{g_{BI}^2\,\ell_{\mathrm{sp}}^4}
        \Bigl[
          \sqrt{\det\!\left(\delta^\mu_\nu
                        + \ell_{\mathrm{sp}}^2\,M^\mu{}_\nu\right)} - 1
          \Bigr],
        \qquad
        M^\mu{}_\nu = F^{a\mu\rho}F^a_{\rho\nu},
        \label{eq:BI-gauge}
      \end{equation}
      which is manifestly symmetric in its matrix argument and reduces to the standard
      BI form in the abelian case.

      \begin{lemma}[Gauge BI reduces to Yang--Mills]
        \label{lem:BI-YM}
        In the limit $\ell_{\mathrm{sp}}^2 M^{\mu}{}_\nu \to 0$:
        \[
          \mathcal{L}_{BI}[A]
          \;=\;
          -\tfrac{1}{4}\mathrm{Tr}(F_{\mu\nu}F^{\mu\nu})
          \;+\;
          O\!\left(\ell_{\mathrm{sp}}^4 \mathrm{Tr}(M)^2\right).
        \]
      \end{lemma}

      \begin{proof}
        Expand $\sqrt{\det(I + \ell^2_{\mathrm{sp}} M)} = 1 + \tfrac{1}{2}\ell^2_{\mathrm{sp}}\mathrm{tr}(M)
        + O(\ell^4_{\mathrm{sp}})$.
        Since $\mathrm{tr}(M^\mu{}_\nu) = F^{a\mu\rho}F^a_{\rho\mu} = -\mathrm{Tr}(F_{\mu\nu}F^{\mu\nu})$,
        the leading term is $-\tfrac{1}{2}\ell^4_{\mathrm{sp}}\,g^{-2}_{BI}\,\mathrm{Tr}(F^2)$,
        matching the Yang--Mills Lagrangian $-\tfrac{1}{4}\mathrm{Tr}(F_{\mu\nu}F^{\mu\nu})$
        with $g_{YM} = g_{BI}$.
        The symmetric argument $M$ avoids the sign ambiguity of $\sqrt{\det(\delta + \ell^2 F)}$
        in the non-abelian case.
      \end{proof}

    \subsection{Joint Eddington--Born--Infeld functional}
      \label{ssec:joint-EBI}

      \begin{definition}[Joint EBI functional]
        \label{def:EBI}
        The joint Eddington--Born--Infeld extension of the projective spectral entropy
        functional is:
        \begin{align}
          S_{EBI}[g,A]
          &=
          \frac{2}{\kappa^2}
          \int_M
          \Bigl[
            \sqrt{\det\!\left(g_{\mu\nu}
                          + \ell_{\mathrm{sp}}^2\,\bigl(R_{\mu\nu} + \kappa F^a_{\mu\rho}F^{a\rho}{}_\nu\bigr)
            \right)}
            - \sqrt{g}
            \Bigr]
          d^4x,
          \label{eq:EBI}
        \end{align}
        where $\kappa^2 = 8\pi G_N$ and the Ricci tensor $R_{\mu\nu}$ is computed from $g$.
      \end{definition}

      \begin{theorem}[EBI recovers joint Einstein--Yang--Mills at leading order]
        \label{thm:EBI}
        In the limit $\ell_{\mathrm{sp}}^2(R_{\mu\nu} + \kappa F^2_{\mu\nu})\to 0$, the
        variation of $S_{EBI}[g,A]$ reproduces the joint system~\eqref{eq:Einstein-YM}--\eqref{eq:YM}
        at leading order.
        The EBI functional is regular at $\ell_{\mathrm{sp}}$ in both sectors and
        satisfies the following boundary conditions:
        \begin{enumerate}
          \item[\textrm{(EBI-1)}] $S_{EBI}[g,A]\to c_{EH} a_2[g] + c_{YM} a_4[g,A]$ as $\ell_{\mathrm{sp}}\to 0$;
          \item[\textrm{(EBI-2)}] $S_{EBI}[g,A]$ is bounded above by $\ell_{\mathrm{sp}}^{-4}\mathrm{Vol}(M)$;
          \item[\textrm{(EBI-3)}] $S_{EBI}$ is a minimal admissible diffeomorphism-invariant completion
          satisfying (EBI-1) and (EBI-2) within the tensorial class in which it depends on
          $g$ and $A$ only through the combination $g_{\mu\nu} + \ell_{\mathrm{sp}}^2(R_{\mu\nu} + \kappa F^2_{\mu\nu})$;
          uniqueness within this class is partial, subject to the residual ambiguity
          in the precise form of $\Xi_{\mu\nu}$ noted in~\cite{Beau2026c}.
        \end{enumerate}
      \end{theorem}

      \begin{proof}
        Expanding the determinant in~\eqref{eq:EBI}:
        \begin{align}
          \sqrt{\det(g + \ell^2_{\mathrm{sp}}\Xi)}
          &= \sqrt{g}
          \Bigl[
            1
            + \tfrac{1}{2}\ell_{\mathrm{sp}}^2 g^{\mu\nu}\Xi_{\mu\nu}
            - \tfrac{1}{4}\ell_{\mathrm{sp}}^4\bigl(
            \tfrac{1}{2}(g^{\mu\nu}\Xi_{\mu\nu})^2
            - g^{\mu\rho}g^{\nu\sigma}\Xi_{\mu\nu}\Xi_{\rho\sigma}
            \bigr)
            + O(\ell_{\mathrm{sp}}^6\Xi^3)
            \Bigr],
          \nonumber
        \end{align}
        where $\Xi_{\mu\nu} = R_{\mu\nu} + \kappa F^a_{\mu\rho}F^{a\rho}{}_\nu$.
        The $\ell_{\mathrm{sp}}^2$ term gives $\kappa^{-2}\ell_{\mathrm{sp}}^2 R\sqrt{g}/2$
        for the Ricci scalar part, matching $c_{EH} a_2[g]$ under the identification
        $\ell_{\mathrm{sp}}^2 = \kappa^2 c_{EH}/2$.
        The gauge contribution at this order is $\ell_{\mathrm{sp}}^2 F^2\sqrt{g}$,
        which matches $c_{YM} a_4$ upon renormalization.
        Property~(EBI-2) follows from the bound $|\det(I + X)| \leq e^{\mathrm{tr}(X)}$ applied to
        $X = \ell_{\mathrm{sp}}^2 g^{-1}\Xi$.
        Uniqueness within the ansatz class is partial; property~(EBI-3) follows by the argument of~\cite{Beau2026c} applied to
        the joint tensor $\Xi_{\mu\nu}$, up to the residual ambiguity in the traceless modification of $R_{\mu\nu}$.
      \end{proof}

      \begin{remark}
        The joint EBI functional~\eqref{eq:EBI} is the Cosmochrony analogue of the
        Bañados--Ferreira--Skordis theory~\cite{Banados2010} and of the Eddington action,
        extended to include gauge curvature.
        The admissibility saturation bound $c_\chi$ provides the physical justification
        for the BI structure: the effective amplitude of any admissible mode is bounded
        above by $c_\chi/\sqrt{\lambda_n}$, which at the field-theory level becomes the
        saturation of the determinant form.
        The residual ambiguity in the precise form of $\Xi_{\mu\nu}$ (whether to use
        $R_{\mu\nu}$ or $R_{\mu\nu} - \tfrac{1}{4}g_{\mu\nu}R$) noted in~\cite{Beau2026c}
        persists in the joint functional and is the primary open point of the completion.
      \end{remark}

  \section{The gauge--gravity hierarchy ratio}
    \label{sec:hierarchy}

    \subsection{The two induced couplings}
      \label{ssec:two-couplings}

      From~\cite{Beau2026h} and~\cite{Beau2026q12}, the two induced couplings of the
      joint functional are given in~\eqref{eq:induced-couplings}.
      Both emerge from the same spectral cutoff $\ell_{\mathrm{sp}}$ and the
      same renormalization group flow; they differ in the UV divergence structure
      of their respective Seeley--DeWitt coefficients ($\Lambda^2$ versus $\log\Lambda$).
      In particular:
      \begin{equation}
        G_N \;\sim\; \frac{16\pi^2}{\dim V}\,\ell_{\mathrm{sp}}^{2}
        \qquad\text{and}\qquad
        g_{YM}^{2} \;\sim\; \frac{12\cdot 16\pi^2}{\log(\Lambda/\mu)}.
        \label{eq:couplings-explicit}
      \end{equation}

    \subsection{The hierarchy ratio}
      \label{ssec:ratio}

      \begin{proposition}[Gauge--gravity hierarchy]
        \label{prop:hierarchy}
        The ratio of the two induced couplings is:
        \begin{equation}
          G_N\,g_{YM}^2
          \;=\;
          \frac{12\,\ell_{\mathrm{sp}}^2}{\dim V}
          \cdot\Bigl[\log\frac{\Lambda}{\mu}\Bigr]^{-1}.
          \label{eq:ratio}
        \end{equation}
        For $\ell_{\mathrm{sp}} < \ell_{\mathrm{Planck}}$ and $\Lambda/\mu\gg 1$:
        \begin{equation}
          G_N\,g_{YM}^2 \;\ll\; 1.
          \label{eq:hierarchy}
        \end{equation}
      \end{proposition}

      \begin{proof}
        Substitute~\eqref{eq:couplings-explicit} and multiply:
        \[
          G_N\,g_{YM}^2
          \;\sim\;
          \frac{16\pi^2\,\ell_{\mathrm{sp}}^2}{\dim V}
          \cdot
          \frac{12\cdot 16\pi^2}{(16\pi^2)^2\,\log(\Lambda/\mu)}
          \;=\;
          \frac{12\,\ell_{\mathrm{sp}}^2}{\dim V\,\log(\Lambda/\mu)}.
        \]
        The numerical factors of $16\pi^2$ cancel between $G_N$ and $g_{YM}^2$.
        With $\ell_{\mathrm{sp}} \sim 10^{-35}$ m and $\log(\Lambda/\mu) \sim \mathcal{O}(10^2)$
        at electroweak scale, and $\dim V \geq 4$ for $G_\Pi \supseteq \mathrm{SU}(2)\times\mathrm{U}(1)$,
        the ratio is suppressed by $\ell_{\mathrm{sp}}^2 \sim 10^{-70}$ m$^2$, confirming~\eqref{eq:hierarchy}.
      \end{proof}

    \subsection{Structural interpretation}
      \label{ssec:structural}

      The gauge--gravity hierarchy~\eqref{eq:hierarchy} is a structural consequence of
      the Seeley--DeWitt order separation between the two sectors:
      \begin{equation}
        \underbrace{a_2 \;\longrightarrow\; \text{gravity},\; G_N^{-1} \sim \ell_{\mathrm{sp}}^{-2}}_{%
          \text{quadratic UV divergence, non-renormalizable}}
        \qquad
        \underbrace{a_4 \;\longrightarrow\; \text{gauge},\; g_{YM}^{-2} \sim \log\Lambda}_{%
          \text{logarithmic UV divergence, renormalizable}}.
        \label{eq:hierarchy-summary}
      \end{equation}
      The single spectral cutoff $\ell_{\mathrm{sp}}$ generates two qualitatively
      different divergence structures.
      The hierarchy $G_N g_{YM}^2 \ll 1$ is therefore a consequence of the spectral
      stratification of the theory and requires no separate fine-tuning or
      anthropic selection.

      \begin{remark}
        The $\dim V$ factor in~\eqref{eq:ratio} provides a mild sensitivity of the
        hierarchy ratio to the gauge group.
        For $G_\Pi = \mathrm{SU}(2)\times\mathrm{U}(1)$ (unconditional~\cite{Beau2026q6a}),
        $\dim V = 4$; for $G_\Pi = \mathrm{SU}(3)\times\mathrm{SU}(2)\times\mathrm{U}(1)$
        (proved unconditionally on the standard graph; $[H\text{-color}]_{\mathrm{pointwise}}$
        established analytically in~\cite{Beau2026a35} via the single-frequency BI fingerprint
        structure, numerical confirmation in~\cite{Beau2026a36}), $\dim V = 12$ in
        the fundamental representation.
        The ratio~\eqref{eq:ratio} is thus three times smaller for the full Standard Model
        gauge group, increasing the hierarchy suppression.
      \end{remark}

  \section{Epistemic status}
    \label{sec:epistemic}

    \subsection{Status of each result}
      \label{ssec:status}

      Table~\ref{tab:status} collects the epistemic status of the main results of this paper.

      \begin{table}[h]
        \centering
        \renewcommand{\arraystretch}{1.3}
        \begin{tabular}{lll}
          \toprule
          Result & Status & Dependence \\
          \midrule
          $\delta_g S_\Pi \ni c_{YM} T^{YM}_{\mu\nu}$ (Lemma~\ref{lem:YM-stress})
          & Theorem-level & Metric-independence of $F_{\mu\nu}$ \\
          Coupled EYM system (Theorem~\ref{thm:EYM})
          & Theorem-level & Lemma~\ref{lem:YM-stress}, \cite{Beau2026h,Beau2026q12} \\
          $a_6$ cross-coupling (Theorem~\ref{thm:a6})
          & Standard $+$ structural & \cite{Vassilevich2003} \\
          EBI completion (Theorem~\ref{thm:EBI})
          & Structural, uniqueness ansatz & \cite{Beau2026c} \\
          $\Xi_{\mu\nu}$ ambiguity in EBI & Open & Tensorial completion \\
          Hierarchy ratio (Proposition~\ref{prop:hierarchy})
          & Structural prediction & \cite{Beau2026h,Beau2026q12} \\
          $G_\Pi = \mathrm{SU}(3)\times\mathrm{SU}(2)\times\mathrm{U}(1)$
          & Proved (unconditional, O31~v1.5);
          $[H\text{-color}]_{\mathrm{pointwise}}$
          via single-frequency BI fingerprint & \cite{Beau2026a35,Beau2026a36} \\
          \bottomrule
        \end{tabular}
        \caption{Epistemic status of the main results.}
        \label{tab:status}
      \end{table}

    \subsection{Analytical status of [H-color]}
      \label{ssec:Hcolor}

      The identification $G_\Pi = \mathrm{SU}(3)\times\mathrm{SU}(2)\times\mathrm{U}(1)$
      is taken as input from the spectral admissibility sub-programme.
      The $\mathrm{SU}(3)$ sector is established unconditionally for the
      colour-adapted Cayley graph in~\cite{Beau2026a35}.
      For the standard Cayley graph, $[H\text{-color}]_{\mathrm{eff}}$
      (equality of capacity exponents in the large-$q$ limit) is proved
      analytically in~\cite{Beau2026a35,Beau2026a36} via the intra-coset
      Vandermonde mechanism and the modulation-frequency argument (O32).
      Furthermore, $[H\text{-color}]_{\mathrm{pointwise}}$ (exact equality at
      finite~$q$) is proved in O31 (Proposition~4.23,~\cite{Beau2026a35}): each
      O12 fingerprint vector carries a single frequency
      $B = c_1 b_1 + c_2 b_2 + c_3 b_3$, and the orbit dilation
      $B \mapsto \omega B$ is a bijection of $\mathbb{Z}/q\mathbb{Z}$, so
      $\sigma_c(n) = \sigma_{\omega c}(n) = \sigma_{\omega^2 c}(n)$ holds exactly.
      All results of the present paper are independent of these results;
      the identification of $G_\Pi$ is an input, and the only remaining
      sensitivity is in the value of $\dim V$ entering the hierarchy ratio.

    \subsection{Open points}
      \label{ssec:open}

      The following open points are carried forward:
      \begin{enumerate}
        \item \textbf{EBI ambiguity.}
        The precise form of the tensor $\Xi_{\mu\nu}$ in the joint EBI functional
        (whether to use $R_{\mu\nu}$, $R_{\mu\nu} - \tfrac{1}{4}g_{\mu\nu}R$,
        or another traceless modification) is not uniquely fixed at the level of
        the present paper.
        This ambiguity is inherited from the gravitational BI completion
        of~\cite{Beau2026c} and affects only subleading terms in the IR expansion.

        \item \textbf{Quantitative $a_6$ coefficients.}
        The numerical values of $\gamma_1, \gamma_2, \ldots$ in~\eqref{eq:a6}
        depend on the specific operator $\mathcal{A}_{g,A}$ and require explicit
        computation for the admissible bundle.
        This is deferred to a companion numerical study.

        \item \textbf{Lorentzian continuation.}
        All spectral derivations are performed in Euclidean signature.
        The Lorentzian continuation of the joint EBI functional is established
        at the level of Q5b~\cite{Beau2026q6b} for the geometric sector;
        its extension to the gauge sector remains an open step.

        \item \textbf{$\mathrm{SU}(3)$ Yang--Mills uniqueness.}
        The analogue of Theorem~\ref{thm:EYM} for the internal structure of
        the $\mathrm{SU}(3)$ sector (colour charge structure, structure constants)
        is identified as an open problem in~\cite{Beau2026a35}.
        This is a separate question from $[H\text{-color}]_{\mathrm{eff}}$,
        which is now proved analytically in~\cite{Beau2026a35,Beau2026a36}.
      \end{enumerate}

      \clearpage
  \section{Conclusion}
    \label{sec:conclusion}

    The main result of this paper is the derivation of the coupled Einstein--Yang--Mills
    system~\eqref{eq:Einstein-YM}--\eqref{eq:YM} as the joint variational problem of the
    projective spectral entropy functional $S_\Pi[g,A]$, completing the synthesis
    announced in~\cite{Beau2026q12}.

    \paragraph{The central structural observation.}
      The joint functional $S_\Pi[g,A] = \tfrac{1}{2}\log\det'\mathcal{A}_{g,A}$
      encodes both gravitational and gauge dynamics in its Seeley--DeWitt expansion:
      \begin{align*}
        a_2 &\;\longrightarrow\; \text{Einstein gravity},\quad
        G_{\mu\nu} = 0 \text{ (in vacuum)},\\
        a_4 &\;\longrightarrow\; \text{Yang--Mills gauge dynamics},\quad
        D_\mu F^{a\mu\nu} = 0,\\
        a_4 \text{ under } \delta_g &\;\longrightarrow\;
        \text{gauge--matter coupling to gravity},\quad
        G_{\mu\nu} = 8\pi G_N T^{YM}_{\mu\nu},\\
        a_6 &\;\longrightarrow\; \text{gauge--gravity cross-coupling},\quad
        R_{\mu\nu\rho\sigma}F^{\mu\nu}F^{\rho\sigma}.
      \end{align*}
      The hierarchy of interactions is identified with a hierarchy of spectral order.
      This identification is the conceptually durable contribution of the present trilogy:
      gravity and gauge dynamics are not unified by enlarging the symmetry group or
      adding fields, but are \emph{stratified} by the Seeley--DeWitt level at which
      the same operator $\mathcal{A}_{g,A}$ responds to variation.
      The different UV behaviours of the two sectors --- quadratic divergence for gravity,
      logarithmic divergence for gauge --- are a direct consequence of this spectral
      stratification, as is the gauge--gravity hierarchy $G_N g_{YM}^2 \ll 1$,
      which is structurally suppressed in $\ell_{\mathrm{sp}}^2$ and $[\log(\Lambda/\mu)]^{-1}$
      without fine-tuning.

    \paragraph{What the programme is.}
      This trilogy (Gravity~3.0~\cite{Beau2026h}, Q12~\cite{Beau2026q12}, Q13) can now
      be stated in one sentence: \emph{the Cosmochrony programme is a spectrally
  stratified theory of emergent interactions, in which the type of a fundamental
  interaction is determined by the Seeley--DeWitt order at which the admissible
  operator responds.}
      The conventional question ``what symmetry unifies gravity and gauge?'' is replaced by:
      ``at what spectral order does the projection respond?''

    \paragraph{Remaining open points.}
      The Eddington--Born--Infeld completion of Section~\ref{sec:EBI} provides a
      unified non-linear regulator at $\ell_{\mathrm{sp}}$, recovering both
      sector-specific completions as limits.
      The $a_6$ cross-coupling of Section~\ref{sec:a6} is suppressed by
      $\ell_{\mathrm{sp}}^2$ and becomes relevant only near $\ell_{\mathrm{sp}}$.
      The residual open points --- EBI tensor ambiguity, $a_6$ numerical coefficients,
      Lorentzian continuation, $\mathrm{SU}(3)$ Yang--Mills uniqueness --- are each
      well-posed and bounded in scope.
      They do not affect the structural hierarchy result~\eqref{eq:ratio} or
      the coupled equations of motion~\eqref{eq:Einstein-YM}--\eqref{eq:YM}.
      The natural continuation of the programme is the treatment of fermionic matter
      and chirality, where the spectral stratification principle predicts the
      existence of a dedicated spectral level carrying spinor structure.

      \clearpage
      \bibliographystyle{plain}
      \bibliography{cosmochrony-bibliography,external-refs}

\end{document}
